---
title: "Dissertation_code"
output: html_document
date: "2023-07-10"
---

-confidence intervals, weekly data, start writing
- acf(nyc$tmax_degrees_fahrenheit, na.action = na.pass)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
library(ismev)
library(tidyverse)
library(lubridate)
library(janitor)
```

#Function 1 - read and clean dataset
```{r}
clean_data <- function(data){
  #cleaning column names for easier referencing
  data <- data %>% clean_names()
  #formatting date column
  data$date <- as_date(data$date, format="%Y-%m-%d")
  print(colSums(is.na(nyc), na.rm= FALSE))
  return(data)
}

```

#Function 2 - Plot monthly trend of area
```{r}
monthly_trend_boxplot <- function(dataset){
  #inserting a month column into dataset
  dataset$month <- format(as.Date(dataset$date, format ="%Y-%m-%d"), "%m")
  #grouping data according to the month
  data_monthly <- dataset %>% group_by(month)
  #creating a boxplot to show monthly temp trends
  boxplot(data_monthly$tmax_degrees_fahrenheit ~ dataset$month,
          xlab = "Month",
          ylab = "Temperature in Farenheit",
          main = "Max Temperature recorded monthly (1869-2023) ")
}
```


#Function 3 - Distribute data into seasons
```{r}
split_data_into_season <- function(data){
  #creating new column to store season
  data$season <- ""
  
  for (m in 1:nrow(data)) {
  month <- as.numeric(format(data$date[m], "%m"))
  if (month %in% c(11,12, 1, 2)) {
    data$season[m] <- "winter"
  } 
  else if (month %in% c(3, 4, 5)) {
    data$season[m] <- "spring"
  } 
  else if (month %in% c(6, 7, 8)) {
    data$season[m] <- "summer"
  } 
  else if (month %in% c(9, 10)) {
    data$season[m] <- "autumn"
  }
  }
  return(data)
}
```

#Function 4 - Covariate Matrix
```{r}
main_matrix <- function(data){
  data$season <- factor(data$season)
  obs <- nrow(data)
  
  data <- data %>%  mutate(week = week(date))
  
  season_matrix <- matrix(0, nrow=obs, ncol= 5)
  
  season_matrix[,1] <- week(data$date)
  season_matrix[,2] <- as.integer(data$season == 'spring')
  season_matrix[,3] <- as.integer(data$season == 'summer')
  season_matrix[,4] <- as.integer(data$season == 'autumn')
  season_matrix[,5] <- as.integer(data$season == 'winter')
  
  return(season_matrix)
}
```

#Function 5 - Plotting monthly trends
```{r}
seasonal_trend_plot <- function(data){
  data %>% 
  filter(year(date) %in% c(1990:2000)) %>% 
  ggplot(aes(x = date,
             y = tmax_degrees_fahrenheit))+
    geom_point()+
      facet_grid(~season)+
    xlab("Year")+
    ylab("Maximum Daily Temperature")
}
```

#Function 6 - Combining daily temp data into weeks
```{r}
distribute_into_summer_weeks <- function(data){
  summer_max <- c()
  summer_data <- data %>% filter (season == "summer")
  
  for (y in 1869:2023){
    year_index_s <- summer_data %>% filter(year(date) == y)
    unique_week_s <- c(unique(week(year_index_s$date)))
    for( w in unique_week_s ){  
      #create block of 1 year with data from that year  
      index_week_s <- which(week(year_index_s$date) == w)
      if (length(year_index_s$tmax_degrees_fahrenheit[index_week_s]) > 0)
        if( sum( is.na( year_index_s$tmax_degrees_fahrenheit[index_week_s] ) ) < 1 ) 
        summer_max <- c( summer_max, max(year_index_s$tmax_degrees_fahrenheit[index_week_s], na.rm=TRUE) )
        else
          next
    }}
  return(summer_max)
}

distribute_into_winter_weeks <- function(data){
  winter_max <- c()
  winter_data <- data %>% filter (season == "winter")
  
  for (y in 1869:2023){
    year_index_w <- winter_data %>% filter(year(date) == y)
    unique_week_w <- c(unique(week(year_index_w$date)))
    for( w in unique_week_w ){  
      #create block of 1 year with data from that year  
      index_week_w <- which(week(year_index_w$date) == w)
      if (length(year_index_w$tmax_degrees_fahrenheit[index_week_w]) > 0)
        if( sum( is.na( year_index_w$tmax_degrees_fahrenheit[index_week_w] ) ) < 1 ) 
          winter_max <- c(winter_max, max(year_index_w$tmax_degrees_fahrenheit[index_week_w], na.rm=TRUE) )
        else
          next
    }}
  return(winter_max)
}
```



#City 1 - New York 
```{r}
nyc <- read.csv('nyc_data.csv', skip=1)

nyc <- clean_data(nyc)
print(head(nyc))

```


```{r}
monthly_trend_boxplot(nyc)
```
#A strong pattern of variation is visible over time across different months in a year, all following a similar trend in temperature. 


##Unnecessary
#Plotting a linear model to look at temperature trend over years 1900-200 in NYC of max temp, min temp, precipitation and snow. 
```{r}
yearly_max <- c()

year <- c(1900:2000)
for( y in 1900:2000 ){
  #create block of 1 year with data from that year
  index_year <- which(year(nyc$date) == y )
  if( sum( is.na( nyc$tmax_degrees_fahrenheit[index_year] ) ) < 60 ) 
    yearly_max <- c( yearly_max, max(nyc$tmax_degrees_fahrenheit[index_year], na.rm=TRUE) )
}

linear_model1 <- lm(yearly_max ~ year)
plot(year,yearly_max)
abline(linear_model1)

```
#A careful look at these data suggests that the pattern of variation has not remained constant over the observation period. There is a discernible increase in the data through time, though the increase seems slighter in more recent years. Just a slight increase can have major effect on the earth's global temperature and a cause of global warming. 



```{r}
#categorising data into nineteenth cetnury (1900-2000)
#cen_20 <-nyc %>% filter(between(date,as_date("1900-01-01"),as_date("2000-01-01")))
```

#Fitting GEV model where variation over the years is modelled as a linear trend, includes non-stationary data
```{r}
ti<- matrix(ncol=1,nrow=101)
ti[,1]=seq(1,101,1)


B <- gev.fit(xdat= yearly_max,ydat=ti, mul=1 )
gev.diag(B)
```


#Gev model fit into stationary data
```{r}
A <- gev.fit(yearly_max)
gev.diag(A)
```

```{r}
deviance <- -2 * (B$nllh-A$nllh)
deviance
```

#Model B is the better fit when comapring the maximising log-likelihood to the chi-square value 3.814. 
#The stationary data model is the better model when comparing the maximized log-likelihood between both the models. 



#Season approach

An alternative, that is especially useful for modeling seasonal changes in threshold exceedance models, is to specify a model with different parameters in each season. Different thresholds might also be appropriate to consider for each season. 
But segregation of seasons itself is an issue. A different situation which may arise is that the extremal behavior of one series is related to that of another variable, referred to as a covariate.

In each case the extreme value parameters can be written in the form
$$\theta(t) = h(X^t\beta)$$,
where $\theta$ denotes either $\mu, \sigma$ or $\xi$, h is a specified function, $\beta$ is a vector of parameters, and X is a model vector. In this context, h is usually referred to as the inverse-link function.

The above equation applied to any or each of the parameters in an extreme value model provides a broad and attractive family for representing non-stationary in extreme value datasets. 

Dividing data into season being spring,summer,autumn and winter. Summer include the month June to August, Winter include the month November to March. Autumn includes August to October. Spring is between March and June.

#Distributing data into seasons
```{r}
nyc <- split_data_into_season(nyc)
```

#Plotting the maximum daily temperature distributed by the seasons for years between 1990 to 2000
```{r}
seasonal_trend_plot(nyc)
```

#creating covariance matrix with seasons
```{r}
nyc_matrix <- main_matrix(nyc)
nyc_matrix[1:10,]
```

#creating blocks of week for summer and winter months
```{r}
summer_max <- distribute_into_summer_weeks(nyc)
summer_max[1:10]
```


#finding thereshold
```{r}
mrl.plot(summer_max, umin = 75, umax=110) 
```


#Stationary GEV model
```{r}
M3 <- gpd.fit(summer_max, threshold = 97)
gpd.diag(M3)
```

#Non-stationary GEV model
```{r}
obs <- nrow(nyc)
years <- length(year(nyc$date))

X <-  matrix(0,nrow = obs, ncol= 2)
X[,1] <- seq(1,years,1)
X[,2] <- X[,1]**2

M4 <- gpd.fit( xdat=summer_max, threshold = 97, ydat=X, sigl = c(1,2) )
gpd.diag(M4)
```
A slight deviations from straight line of residual quantile plot can be observed, suggesting the model is a good fit on the data.

```{r}
#calculating the deviance statistic
deviance_2 <- -2 * (M4$nllh-M3$nllh)
deviance_2 

```
#The deviance value is larger than the chi-square value 3.841 which suggests the presence of a relationship between year and temperature of summer months. The quadratic trend component explains a substantial amount of the variation in the data and is likely to be a geniune effect in the seasonal approach to daily temperature data. (Coles,2001)

```{r}
M4.2 <- gpd.fit( xdat=summer_max, threshold = 97, ydat=nyc_matrix, sigl = c(1,3) )
gpd.diag(M4.2)
```
```{r}
deviance_2.2 <- -2 * (M4.2$nllh-M3$nllh)
deviance_2.2
```




#Excess plot: In winter there seem to be less value of maximum temperature recorded above 60 Farenheit. Threshold model is fit onto this excess data.
```{r}
excess <- nyc %>% 
  filter(season == "winter") %>% 
  filter(tmax_degrees_fahrenheit > 60)
plot(excess$tmax_degrees_fahrenheit)
```


#finding thereshold
```{r}
winter_max <- distribute_into_winter_weeks(nyc)
winter_max[1:10]

mrl.plot(winter_max, umin = 35, umax=90)
```


#Stationary GEV model
```{r}
M5 <- gpd.fit(winter_max, threshold = 72)
gpd.diag(M5)
```

#Non-stationary GEV model
```{r}
obs <- nrow(nyc)
years <- length(year(nyc$date))

X <-  matrix(0,nrow = obs, ncol= 1)
X[,1] <- seq(1,years,1)


M6 <- gpd.fit( xdat=winter_max, threshold = 72, ydat=X, sigl = c(1) )
gpd.diag(M6)
```
deviations from straight line of residual quantile plot suggests areas of improvement in the model.


```{r}
#calculating the deviance statistic
deviance_3 <- -2 * (M6$nllh-M5$nllh)
deviance_3 

```

#fitting covariance matrix with year and season as the parameters
```{r}
M6.2 <- gpd.fit( xdat=winter_max, threshold =72, ydat=nyc_matrix, sigl = c(1,5) )
gpd.diag(M6.2)
```
```{r}
deviance_3.2 <- -2 * (M6.2$nllh-M5$nllh)
deviance_3.2 
```



#City 2 - Atlantic City

```{r}
atlantic <- read.csv('atlanti.city_data.csv', skip = 1)

atlantic <- clean_data(atlantic)
print(head(atlantic))
```


```{r}
monthly_trend_boxplot(atlantic)
```


```{r}
#GEV model for stationary data
A <- gev.fit(yearly_max)
gev.diag(A)

#Fitting GEV model where variation over the years is modelled as a linear trend, includes non-stationary data

ti<- matrix(ncol=1,nrow=101)
ti[,1]=seq(1,101,1)

B <- gev.fit(xdat= yearly_max,ydat=ti, mul=1 )
gev.diag(B)

#deviance
deviance <- -2 * (B$nllh-A$nllh)
deviance 

```


#Distributing into seasons
```{r}
atlantic <- split_data_into_season(atlantic)
```

```{r}
seasonal_trend_plot(atlantic)
```

```{r}
atlantic_matrix <- main_matrix(atlantic)
atlantic_matrix[1:10,]
```













